Alone Together: Compositional Reasoning and Inference for Weak Isolation

### 2017-10-22 inference.tex

The state transformer rules, like the earlier RG rules, closely follow
the corresponding reduction rules in Fig.~\ref{fig:txnimp}, except
that their language of expression is $\SL$. For instance, while the
reduction rule for \C{UPDATE} declaratively specifies the set of
updated records, the state transformer rule uses $\SL$'s bind
operation to \emph{compute} the set. Other SQL rules do likewise. The
rules for \C{LET} binders, conditionals, and sequences compose the
effects inferred for their subcommands. Thus, the effect of a sequence
of commands $c_1;c_2$ is the union of effects $F_1$ and $F_2$ of $c_1$
and $c_2$, respectively, except that $F_2$ is applied to a state
($\stl$) to which $F_1$ has already been applied, reflecting their
order of reduction. The inference rule for \C{FOREACH} takes advantage
of the $\SL$'s bind operator to lift the effect inferred for the loop
body to the level of the loop. Since records added to $\stl$ in each
iteration of \C{FOREACH} are independent of the previous iteration
(recall local context independence; Sec. ~\ref{sec:opsem}), sequential
composition of the effects of different iterations is same as their
parallel composition. Since the loop body is executed once per each
$z\in x$, the effect of the the loop is the union of effects ($\F$)
for all $z\in x$, all applied to the same state ($\stl$ and $\stg$).
That is, $\F_{loop}(\stl,\stg) = \bigcup_{z\in
x}\F_{body}(\stl,\stg)$. From the definition of the set monad's bind
operator, $\F_{loop}(\stl,\stg) = x \bind (\lambda
z.~F_{body}(\stl,\stg))$, which mirrors the definition of the rule.
This development mirrors ideas found in type checkers for certain
dependent type systems~\cite{KJ14}.

@@

The proofs for \C{SELECT}, sequencing, and conditionals use use the
inductive hypothesis to infer the RG-judgments present in the
premises of their corresponding RG-rules. The interesting case is the
\C{FOREACH} statement, for which we use the loop invariant
$\psi(\stl, \stg) \Leftrightarrow \stl = s \cup (y \bind (\lambda
z.~\F(s, \stg)))$, (where assuming that $c$ is the body of the loop,
$c \elabsto \F$) to prove all the premises of
\rulelabel{RG-Foreach}. Using the same notation as the rule
$\rulelabel{RG-Foreach}$, $y$ refers to the records already processed
in previous iterations of the loop, while $z$ refers to the record
being processed in the current iteration.  At the beginning of the
loop $[\phi/y]\psi(\stl, \stg)$ just reduces to $\stl = s$ which is
implied by the pre-condition $P$. From the inductive hypothesis, we
can infer that each iteration corresponds to the application of
$F$. Since all iterations are assumed to be independent of each other,
and $z$ is bound to a record in $x$ for each iteration, we conclude
that at the end of every iteration, the loop invariant $[y \cup
  \{z\}/y]\psi$ will be satisfied.

@@

Theorem~\ref{thm:inference-sound} lets us replace the local judgment
of the \rulelabel{RG-Txn} rule (Fig.~\ref{fig:rg-rules}) by a state
transformer inference judgment. The soundness of a transaction's
guarantee can now be established w.r.t the effect $\F$ of the body.
The \rulelabel{RG-Txn} rule so updated is shown below:
\begin{smathpar}
\begin{array}{c}
\RULE
{
   \stable(R,\I)\spc
   \stable(R,I)\\
   \R_e = R \backslash \I_e \spc \R_c = R \backslash \I_c \spc  
   c \elabsto \F \spc \stable(\R_c,\F)\\
   \forall \stl,\stg.~ \stl = F(\emptyset,\stg) \Rightarrow 
   G(\stg, \stl\gg\stg)\spc
   \forall \stg,\stg'.~I(\stg) \wedge G(\stg,\stg') \Rightarrow I(\stg')\\
}
{
  \rg{I,R}{\ctxn{i}{\I}{c}}{G,I}
}
\end{array}
\end{smathpar}
% \begin{smathpar}
% \begin{array}{c}
% \RULE
% {
%   \stable(R,\I)\spc
%   \stable(R,I)\spc
%   P(\stl,\stg) \Leftrightarrow \stl=\emptyset \wedge I(\stg)\\
%   \R_e = R \backslash \I_e \spc \R_c = R \backslash \I_c \spc 
%    \R_e \vdash \rg{P}{c}{Q} \spc \stable(\R_c,Q) \\ 
%   \forall \stl,\stg.~ Q(\stl,\stg) \Rightarrow 
%     G(\stg, \stl \rhd \stg)\spc
%   \forall \stg,\stg'.~I(\stg) \wedge G(\stg,\stg') \Rightarrow I(\stg')\\
% }
% {
%   \rg{I,R}{\ctxn{i}{\I}{c}}{G,I}
% }
% \end{array}
% \end{smathpar}
Automating the application of the \rulelabel{RG-Txn} rule for a
transaction requires automating the multiple implication checks in
the premise. While $R$, $G$, $\I$ and $I$ are formulas in
first-order logic (FOL) with a relatively simple structure, $\F$
is an expression in the set language $\SL$
(Fig.~\ref{fig:logic-syntax}) with a possibly complex structure.
Fortunately, however, there exists a semantics-preserving translation
from $\SL$ to a restricted subset of first-order logic (FOL) that
lends itself to automatic reasoning.

>>>>

The state transformer rules, like the earlier RG rules, closely follow
the corresponding reduction rules in Fig.~\ref{fig:txnimp}, except
that their language of expression is $\SL$. For instance, while the
reduction rule for \C{UPDATE} declaratively specifies the set of
updated records, the state transformer rule uses $\SL$'s bind
operation to \emph{compute} the set. Other SQL rules do likewise. The
rules for \C{LET} binders, conditionals, and sequences compose the
effects inferred for their subcommands. Thus, the effect of a sequence
of commands $c_1;c_2$ is the union of effects $\F_1$ and $\F_2$ of $c_1$
and $c_2$, respectively, except that $\F_2$ is applied to a state
($\stl$) to which $\F_1$ has already been applied, reflecting their
order of reduction. The inference rule for \C{FOREACH} takes advantage
of the $\SL$'s bind operator to lift the effect inferred for the loop
body to the level of the loop. Since records added to $\stl$ in each
iteration of \C{FOREACH} are independent of the previous iteration
(recall local context independence; Sec. ~\ref{sec:opsem}), sequential
composition of the effects of different iterations is same as their
parallel composition. Since the loop body is executed once per each
$z\in x$, the effect of the the loop is the union of effects ($\F$)
for all $z\in x$, all applied to the same state ($\stl$ and $\stg$).
That is, $\F_{loop}(\stl,\stg) = \bigcup_{z\in
x}\F_{body}(\stl,\stg)$. From the definition of the set monad's bind
operator, $\F_{loop}(\stl,\stg) = x \bind (\lambda
z.~F_{body}(\stl,\stg))$, which mirrors the definition of the rule.
%This development mirrors ideas found in type checkers for certain
%dependent type systems~\cite{KJ14}.

@@

The proofs for \C{SELECT}, sequencing, and conditionals use use the
inductive hypothesis to infer the RG-judgments present in the
premises of their corresponding RG-rules. The interesting case is the
\C{FOREACH} statement, for which we use the loop invariant
$\psi(\stl, \stg) \Leftrightarrow \stl = s \cup (y \bind (\lambda
z.~\F(s, \stg)))$, (where assuming that $c$ is the body of the loop,
$c \elabsto \F$) to prove all the premises of
\rulelabel{RG-Foreach}. Using the same notation as the rule
$\rulelabel{RG-Foreach}$, $y$ refers to the records already processed
in previous iterations of the loop, while $z$ refers to the record
being processed in the current iteration.  At the beginning of the
loop $[\phi/y]\psi(\stl, \stg)$ just reduces to $\stl = s$ which is
implied by the pre-condition $P$. From the inductive hypothesis, we
can infer that each iteration corresponds to the application of
$\F$. Since all iterations are assumed to be independent of each other,
and $z$ is bound to a record in $x$ for each iteration, we conclude
that at the end of every iteration, the loop invariant $[y \cup
  \{z\}/y]\psi$ will be satisfied.

@@

Theorem~\ref{thm:inference-sound} lets us replace the local judgment
of the \rulelabel{RG-Txn} rule (Fig.~\ref{fig:rg-rules}) by a state
transformer inference judgment. The soundness of a transaction's
guarantee can now be established w.r.t the effect $\F$ of the body.
The \rulelabel{RG-Txn} rule so updated is shown below:
\begin{smathpar}
\begin{array}{c}
\RULE
{
   \stable(R,\I)\spc
   \stable(R,I)\\
   \R_e = R \backslash \I_e \spc \R_c = R \backslash \I_c \spc  
   c \Longrightarrow_{\langle i,\R_e,I \rangle}\F \spc \stable(\R_c,\F)\\
   \forall \stl,\stg.~ \stl = F(\emptyset,\stg) \Rightarrow 
   G(\stg, \stl\gg\stg)\spc
   \forall \stg,\stg'.~I(\stg) \wedge G(\stg,\stg') \Rightarrow I(\stg')\\
}
{
  \rg{I,R}{\ctxn{i}{\I}{c}}{G,I}
}
\end{array}
\end{smathpar}
% \begin{smathpar}
% \begin{array}{c}
% \RULE
% {
%   \stable(R,\I)\spc
%   \stable(R,I)\spc
%   P(\stl,\stg) \Leftrightarrow \stl=\emptyset \wedge I(\stg)\\
%   \R_e = R \backslash \I_e \spc \R_c = R \backslash \I_c \spc 
%    \R_e \vdash \rg{P}{c}{Q} \spc \stable(\R_c,Q) \\ 
%   \forall \stl,\stg.~ Q(\stl,\stg) \Rightarrow 
%     G(\stg, \stl \rhd \stg)\spc
%   \forall \stg,\stg'.~I(\stg) \wedge G(\stg,\stg') \Rightarrow I(\stg')\\
% }
% {
%   \rg{I,R}{\ctxn{i}{\I}{c}}{G,I}
% }
% \end{array}
% \end{smathpar}
Automating the application of the \rulelabel{RG-Txn} rule for a
transaction requires automating the multiple implication checks in
the premise. While $R$, $G$, $\I$ and $I$ are formulas in
first-order logic (FOL) with a relatively simple structure, $\F$
is an expression in the set language $\SL$
(Fig.~\ref{fig:logic-syntax}) with a possibly complex structure.
Fortunately, however, there exists a semantics-preserving translation
from $\SL$ to a restricted subset of first-order logic (FOL) that
lends itself to automatic reasoning.

### 2017-10-22 reasoning.tex
Minor edits.

The stability condition on $\I$ guarantees that an interference from
$R^*$ is admissible only if the interference due to each individual
$R$ step is admissible. In other words, it makes isolation-constrained
$R^*$ relation equal to the transitive closure of the
isolation-constrained $R$ relation. We call $R$ constrained by
isolation $\I$ as $R$ modulo $\I$ ($R\backslash \I$; written shortly
as $\R$), which is the following ternary relation:
\begin{smathpar}
  \begin{array}{lcl}
    (R \backslash \I)(\stl, \stg, \stg') & \Leftrightarrow & R(\stg,
      \stg') \wedge \I(\stl, \stg, \stg')\\
  \end{array}
\end{smathpar}
It is now enough to prove the stability of an RG assertion $\phi$
w.r.t $R\backslash \I$:
\begin{smathpar}
  \begin{array}{lcl}
    \stable((R \backslash \I), \phi) & \Leftrightarrow & \forall
      \stl,\stg,\stg'.~ \phi(\stl,\stg) \wedge (R \backslash \I)(\stl,\stg,\stg') \Rightarrow \phi(\stl,\stg')\\
  \end{array}
\end{smathpar}
This condition often significantly simplifies the form of $R
\backslash \I$ irrespective of $R$. For example, when a transaction is
executed against a snapshot of the database (i.e. $\mathbb{I}_{ss}$),
$R \backslash \I_{ss}$ will be the identity function, since any
non-trivial interference will violate the $\stg' = \stg$ condition
imposed by $\I_{ss}$.

@@

\noindent Thus, if $\rg{I,R}{c}{G,I}$ is a valid RG judgment, then (a)
every interleaved multi-step reduction of $c$ preserves the database
integrity constraint (consistency condition) $I$, and (b) the effect
that every transaction in $c$ has on the database state is captured by
$G$.
\noindent We can now assert the soundness of the RG judgments in
Fig.~\ref{fig:rg-rules} as follows:

>>>>

The stability condition on $\I$ guarantees that an interference from
$R^*$ is admissible only if the interference due to each individual
$R$ step is admissible. In other words, it makes isolation-constrained
$R^*$ relation equal to the transitive closure of the
isolation-constrained $R$ relation. We call $R$ constrained by
isolation $\I$ as $R$ modulo $\I$ ($R\backslash \I$; written equivalently
as $\R$), which is the following ternary relation:
\begin{smathpar}
  \begin{array}{lcl}
    (R \backslash \I)(\stl, \stg, \stg') & \Leftrightarrow & R(\stg,
      \stg') \wedge \I(\stl, \stg, \stg')\\
  \end{array}
\end{smathpar}
It is now enough to prove the stability of an RG assertion $\phi$
w.r.t $R\backslash \I$:
\begin{smathpar}
  \begin{array}{lcl}
    \stable((R \backslash \I), \phi) & \Leftrightarrow & \forall
      \stl,\stg,\stg'.~ \phi(\stl,\stg) \wedge (R \backslash \I)(\stl,\stg,\stg') \Rightarrow \phi(\stl,\stg')\\
  \end{array}
\end{smathpar}
This condition often significantly simplifies the form of $R
\backslash \I$ irrespective of $R$. For example, when a transaction is
executed against a snapshot of the database (i.e. $\mathbb{I}_{ss}$),
$R \backslash \I_{ss}$ will be the identity function, since any
non-trivial interference will violate the $\stg' = \stg$ condition
imposed by $\I_{ss}$.

@@

\noindent Thus, if $\rg{I,R}{c}{G,I}$ is a valid RG judgment, then (a)
every interleaved multi-step reduction of $c$ preserves the database
integrity constraint (consistency condition) $I$, and (b) the effect
that every transaction in $c$ has on the database state is captured by
$G$.
\noindent We can now assert the soundness of the RG judgments in
Fig.~\ref{fig:rg-rules} as follows\footnote{Full proofs for theorems
in this paper can be found in the accompanying tech report}:
